Megan's code for Isoviz

### 2023-04-28 temp_code/MS_isoviz_code.Rmd
MS update grna prediction

Inputs leafcutter clustering output and outputs formatted table
```{r}
# This is data from our A375 screen cells and is the most accurate for our work
# now this includes constitutive exons which have a single junction per cluster
# required 30 total reads per cluster and junction to be 10% of total
clusters <- read_tsv("~/isoviz/data-raw/a375_screen_cells_perind.constcounts.gz", col_names = FALSE, skip = 1) # 59,783
clusters %<>% 
  separate(X1, into = c("coords", "counts"), sep = " ") %>% 
  separate(counts, into = c("junc.counts", "cluster.counts"), sep = "/", convert = TRUE) %>%
  separate(coords, into = c("chr", "start", "end", "name"), sep = ":", convert = TRUE) %>% 
  separate(name, into = c("same", "cluster.n", "extra"), sep = "_") %>% 
  select(-same, -extra)

@@

# Consider all junction counts and use gencode annotations to define unique versus common
```{r}
clusters <- read_tsv("~/isoviz/data-raw/hesc_all_junctions_lfc_perind.constcounts.gz", col_names = FALSE, skip = 1)
clusters %<>% 
  separate(X1, into = c("coords", "counts"), sep = " ") %>% 
  separate(counts, into = c("junc.counts", "cluster.counts"), sep = "/", convert = TRUE) %>%
  separate(coords, into = c("chr", "start", "end", "name"), sep = ":", convert = TRUE) %>% 
  separate(name, into = c("same", "cluster.n", "extra"), sep = "_") %>% 
  select(-same, -extra)

@@

@@

Function: Map junction to isoform
```{r}
load(file = "~/isoviz/data/iso_intron_data.rda")
#load(file = "~/isoviz/data/a375_screen_lfc_clusters.rda")
#load(file = "~/isoviz/data/HEK293_lfc_clusters.rda")

# is it better to load rda within a function?
# map junctions function- for now, it isn't using cell type but eventually we should change this
map_junction = function(gene = "ENSG00000001167.15", cell_type = "HEK293"){

  # filter for gene, for now, also filter for protein coding transcripts- can set this as a parameter later
  gene_iso_data = iso_intron_data %>% filter(grepl(gene, gene_id), transcript_type == "protein_coding") # 17 for NFYA

  # join iso data with leafcutter data and mark any unannotated junctions
  # consider cell type
  lfc_clusters = paste0(cell_type, "_lfc_clusters")
  load(file = paste0("~/isoviz/data/", lfc_clusters, ".rda"))
  const_junc_list = clusters_table %>% 
    group_by(cluster.n) %>% 
    tally() %>% filter(n == 1) %>% ungroup()
  clusters_table %<>% mutate(junc.type = ifelse(cluster.n %in% const_junc_list$cluster.n, "Common", "Unique"))
  gene_cluster = gene_iso_data %>% 
    left_join(clusters_table, by = c("chr", "intron_starts" = "start", "intron_ends" = "end")) %>%
    filter(!is.na(cluster.n))

  gene = unique(gene_cluster$gene_id)
  name = unique(gene_cluster$gene_name)
  gene_strand = unique(gene_cluster$strand)
  cluster_ids = unique(gene_cluster$cluster.n)

  output = clusters_table %>% filter(cluster.n %in% cluster_ids) %>% group_by(cluster.n) %>% mutate(junc.per.cluster = n()) %>% 
    left_join(gene_iso_data, by = c("chr", "start" = "intron_starts", "end" = "intron_ends")) %>% 
    mutate(junc.usage = round((junc.counts/cluster.counts)*100, digits = 0), 
           transcript_name = ifelse(is.na(transcript_name), paste0(cluster.n, "_Unannotated"), transcript_name),
           trans_id = ifelse(is.na(trans_id), paste0(cluster.n, "_Unannotated"), trans_id),
           gene_id = ifelse(is.na(gene_id), gene, gene_id),
           gene_name = ifelse(is.na(gene_name), name, gene_name),
           strand = ifelse(is.na(strand), gene_strand, strand), cell_line = paste0(cell_type))

#to_plot = map_junction("ENSG00000001167.15") # NFYA
#to_plot = map_junction("ENSG00000100320") # RBFOX2, alt start is missing because it's intron doesnt overlap another and counts are only 8 so included in consitutive either

# can we ask if all junctions of an isoform are represented to filter isoforms?
# can we create pseudo regtools output based on all possible junctions and run that through leafcutter to get true common versus unique

@@

# Outdated
```{r}
load(file='~/isoviz/data/iso_data_complete.rda')

# plot isoform plus leafcutter clusters
plot_junction_to_isoform = function(df, include_common_juncs = FALSE){

  # I am copying the plot_isoforms function here, figure out how to integrate these later
  gene = unique(df$gene_id)
  gene_trans = which(str_detect(iso_data_complete$gene_id, gene))
  gene_data = iso_data_complete[gene_trans,]

  # filter out common junctions for plotting (default)
  if(include_common_juncs == FALSE){
    df %<>% filter(junc.type == "Unique")
  }

  # get info
  name = unique(df$gene_name)
  transcripts = unique(df$transcript_name)

  # filter for transcripts that have junction level data to support
  gene_data %<>% filter(transcript_name %in% transcripts)

  # make isoform plot
  p1 = ggplot() +
    geom_segment(aes(x = to_plot_ordered$start, y = to_plot_ordered$trans_order,
                     xend = to_plot_ordered$seg_end, yend = to_plot_ordered$trans_order)) +
    geom_rect(data = to_plot_ordered, mapping = aes(xmin = blockstarts, xmax = blockends, ymin = trans_order - 0.3, ymax = trans_order + 0.3)) +
    xlim(min_start, max_end) + ylim(0,y_max) +
    geom_text(aes(x = Inf, y = to_plot_ordered$trans_order, hjust = 0, label = to_plot_ordered$transcript_name), size = 3) +
    theme_bw() + ylab("") + xlab("") +
    theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(), 
          axis.text.y=element_blank(), axis.ticks.y=element_blank(), legend.position = "top", 
          plot.title = element_text(hjust = 0.5), plot.margin = margin(0.1,1,0,0.1, "in")) +
    ggtitle(paste0(name, " Junction to Isoform Map (" , to_plot_ordered$strand[1], " strand )")) + 
    coord_cartesian(xlim = c(min_start, max_end), clip = 'off')

  # new code to plot leafcutter cluster info underneath isoform level data
  # aggregate isoform information
  text = df %>% separate(col = transcript_name, into = c("name", "transcript")) %>% group_by(cluster.n, start, end) %>% 
    arrange(cluster.n, start, end, transcript) %>%
    mutate(isoforms = paste0(transcript, collapse = ",")) %>%
    select(1:5, junc.usage, isoforms) %>% distinct() %>% ungroup() %>% 
    mutate(text_plot = paste0(junc.counts, " reads (", junc.usage, "%) : ", isoforms))

  clusters = unique(df$cluster.n)
  introns = df %>% select(-gene_id, -trans_id, -gene_name, -transcript_name) %>% 
    distinct() %>% arrange(cluster.n, start)
  introns$junc.order = 1:nrow(introns)
  introns %<>% left_join(text)
  p2 = ggplot() + 
    geom_rect(data = introns, mapping = aes(xmin = start, xmax = end, ymin = junc.order - 0.2, ymax = junc.order + 0.2, fill = cluster.n)) + 
    geom_text(data = introns, aes(x = end, y = junc.order, label = text_plot), hjust = 0, size = 3) +
    xlim(min_start, max_end) + theme_bw() + 
    xlab("Hg38 Genomic Position (bp)") + ylab("") + 
    theme(axis.text.y=element_blank(), axis.ticks.y=element_blank(), legend.position = "none", plot.margin = margin(0,1,0,0.1, "in")) +
    coord_cartesian(xlim = c(min_start, max_end))

  # plot dimenstions based on number of isoforms and junctions
  t = (n + nrow(introns))
  h = t/2
  p1_p2 = plot_grid(p1, p2, ncol = 1, align = "v", rel_heights = c(n*0.8, nrow(introns)))
  #ggsave("~/isoviz/plots/rbfox2_isoforms.pdf", width = 8, height = h)
  return(p1_p2)
}

#plot_junction_to_isoform(to_plot)
#ggsave("~/isoviz/plots/rbfox2_isoforms.pdf", width = 8, height = 8.5)
```

@@

# plot isoform plus leafcutter clusters
# add a parameter to only plot a specific junction
plot_all_junction_to_isoform = function(gene = "ENSG00000001167.15", cell_type = "hESC", junction_usage = 5, include_common_juncs = FALSE, filter_expressed_isoforms = TRUE){

@@

  # gets rid of gencode partially unique junctions that are common in this cell type (100% usage) 
  if(filter_expressed_isoforms == TRUE){
    remove_junc = introns %>% filter(junc.usage & isoforms %in% expressed_isoforms$isoforms)
    introns %<>% filter(!junc_id %in% remove_junc$junc_id)
  }

@@

@@

# need to read in count information a different way- this is just a quick and dirty method
# write a function to get expressed junctions/isoforms and call that within functions?
# some of this is copied from map_all_junctions so should make this a function
junctions = c("junc178149", "junc178148")
junc_list = junctions
get_guide_predictions = function(junc_list, cell_type = "hESC", guides_per_junction = 2){
  load(file = "~/isoviz/data/gencode_predictions.rda")
  filtered_guides = gencode_predictions %>% filter(junc_id %in% junc_list)
  gene_name = unique(filtered_guides$gene_name)
  gene = unique(filtered_guides$gene_id)

@@

  # make the isoforms shorthand- probably should change in original file
  output = top_guides %>% left_join(all_gene_clusters) %>%
    select(Cluster = cluster_idx, JuncID = junc_id, Category = junction_category, Counts = junc.counts, Usage = junc.usage,
           Isoforms = transcript_isoforms, GuideID = guide_id, GuideSeq = guide_sequence, Prediction = predicted_lfc)

  # output table
  # don't add sig digits
  fun <- function(x) {formatC(x, format = "f", digits = 0)}

@@

```{r}
# NFYA
to_plot = map_junction("ENSG00000001167.15", cell_type = "A375") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/nfya_isoforms.pdf", width = 8, height = 4)

to_plot = map_junction("ENSG00000001167.15", cell_type = "HEK293") 
plot_junction_to_isoform(to_plot)
# for height, count rows and divide by 2
ggsave("~/isoviz/plots/nfya_isoforms_hek293.pdf", width = 8, height = 2)

to_plot = map_junction("ENSG00000001167.15", cell_type = "hESC_med_string") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/nfya_isoforms_hESC.pdf", width = 8, height = 5)

# RBFOX2
to_plot = map_junction("ENSG00000100320", cell_type = "A375") # alt start is missing because it's intron doesnt overlap another and counts are only 8 so included in consitutive either
plot_junction_to_isoform(to_plot)

to_plot = map_junction("ENSG00000100320", cell_type = "HEK293") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/rbfox2_isoforms_hek293.pdf", width = 8, height = 9)

to_plot = map_junction("ENSG00000100320", cell_type = "hESC_med_string") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/rbfox2_isoforms_hESC.pdf", width = 8, height = 8.5)

# PTBP1
to_plot = map_junction("ENSG00000011304", cell_type = "hESC_med_string") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/ptbp1_isoforms_hESC.pdf", width = 8, height = 8.5)

to_plot = map_junction("ENSG00000011304", cell_type = "HEK293") 
plot_junction_to_isoform(to_plot) # no unique junctions here, fix code

to_plot = map_junction("ENSG00000011304", cell_type = "A375") 
plot_junction_to_isoform(to_plot) # no unique junctions here

# MKNK2
to_plot = map_junction("ENSG00000099875", cell_type = "hESC_med_string") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/mknk2_isoforms_hESC.pdf", width = 8, height = 4)

to_plot = map_junction("ENSG00000099875", cell_type = "HEK293") 
plot_junction_to_isoform(to_plot)

to_plot = map_junction("ENSG00000099875", cell_type = "A375") 
plot_junction_to_isoform(to_plot)

# HNRNPA1
to_plot = map_junction("ENSG00000135486", cell_type = "hESC_med_string") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/hnrnpa1_isoforms_hESC.pdf", width = 8, height = 5.5)

# HNRNPK
to_plot = map_junction("ENSG00000165119", cell_type = "hESC_med_string") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/hnrnpk_isoforms_hESC.pdf", width = 16, height = 6)

# KHSRP- ENSG00000088247
to_plot = map_junction("ENSG00000088247", cell_type = "hESC_med_string") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/khsrp_isoforms_hESC.pdf", width = 16, height = 6)

# MATR3
to_plot = map_junction("ENSG00000015479", cell_type = "hESC_med_string") 
plot_junction_to_isoform(to_plot)
ggsave("~/isoviz/plots/matr3_isoforms_hESC.pdf", width = 16, height = 6)

>>>>

Inputs leafcutter clustering output and outputs formatted table
This is old since we use minicutter now
```{r}
# This is data from our A375 screen cells and is the most accurate for our work
# now this includes constitutive exons which have a single junction per cluster
# required 30 total reads per cluster and junction to be 10% of total
clusters <- read_tsv("~/isoviz/data-raw/a375_screen_cells_perind.constcounts.gz", col_names = FALSE, skip = 1) # 59,783
clusters %<>% 
  separate(X1, into = c("coords", "counts"), sep = " ") %>% 
  separate(counts, into = c("junc.counts", "cluster.counts"), sep = "/", convert = TRUE) %>%
  separate(coords, into = c("chr", "start", "end", "name"), sep = ":", convert = TRUE) %>% 
  separate(name, into = c("same", "cluster.n", "extra"), sep = "_") %>% 
  select(-same, -extra)

@@

# Consider all junction counts and use gencode annotations to define unique versus common
# this is also irrelevant now since we have minicutter
```{r}
clusters <- read_tsv("~/isoviz/data-raw/hesc_all_junctions_lfc_perind.constcounts.gz", col_names = FALSE, skip = 1)
clusters %<>% 
  separate(X1, into = c("coords", "counts"), sep = " ") %>% 
  separate(counts, into = c("junc.counts", "cluster.counts"), sep = "/", convert = TRUE) %>%
  separate(coords, into = c("chr", "start", "end", "name"), sep = ":", convert = TRUE) %>% 
  separate(name, into = c("same", "cluster.n", "extra"), sep = "_") %>% 
  select(-same, -extra)

@@

### Potential function #1: ###

@@

@@

@@

# plot isoform plus leafcutter clusters
# add a parameter to only plot a specific junction
plot_all_junction_to_isoform = function(gene = "ENSG00000099875", cell_type = "hESC", junction_usage = 5, include_common_juncs = FALSE, filter_expressed_isoforms = TRUE, filter_all_100percent = FALSE){

@@

  # gets rid of gencode partially unique junctions that are common in this cell type (100% usage) 
  if(filter_expressed_isoforms == TRUE & filter_all_100percent == FALSE){
    remove_junc = introns %>% filter(junc.usage == "100" & isoforms %in% expressed_isoforms$isoforms)
    introns %<>% filter(!junc_id %in% remove_junc$junc_id)
  }else{
    introns %<>% filter(junc.usage != "100")
  }

@@

plot_all_junction_to_isoform("ENSG00000099875", cell_type = "hESC", filter_all_100percent = TRUE) # MKNK2
ggsave("~/isoviz/plots/mknk2_filtered.png", width = 8, height = 3)

@@

# need to read in count information a different way- this is just a quick and dirty method
# write a function to get expressed junctions/isoforms and call that within functions?
# some of this is copied from map_all_junctions so should make this a function
junctions = c("junc178149", "junc178148")
junctions = c("junc128207", "junc128208")

junc_list = junctions
get_guide_predictions = function(junc_list, cell_type = "hESC", guides_per_junction = 2){
  load(file = "~/isoviz/data/gencode_predictions.rda")
  filtered_guides = gencode_predictions %>% filter(junc_id %in% junc_list)
  gene_name = unique(filtered_guides$gene_name)
  gene = unique(filtered_guides$gene_id)

@@

  # make the isoforms shorthand- probably should change in original file
  output = top_guides %>% left_join(all_gene_clusters) %>%
    select(Cluster = cluster_idx, JuncID = junc_id, Category = junction_category, Counts = junc.counts, Usage = junc.usage,
          GuideSeq = guide_sequence, Prediction = predicted_lfc)
  #select(Cluster = cluster_idx, JuncID = junc_id, Category = junction_category, Counts = junc.counts, Usage = junc.usage,Isoforms = transcript_isoforms, GuideID = guide_id, GuideSeq = guide_sequence, Prediction = predicted_lfc)
  # output table
  # don't add sig digits
  fun <- function(x) {formatC(x, format = "f", digits = 0)}

@@
